---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# polyafit

<!-- badges: start -->
<!-- badges: end -->


`polyafit` is an R package that provides tools to fit count data to a
[Dirichlet-multinomial distribution](https://en.wikipedia.org/wiki/Dirichlet-multinomial_distribution).

This distribution is sometimes called the multivariate Pólya distribution, from
which the package name is derived. The package was created specifically for
16S rRNA marker gene sequence data from the human microbiome, but in principle
could be applied to other types of data.

## Installation

You can install polyafit from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kylebittinger/polyafit")
```

## Example

```{r, echo=FALSE, message=FALSE}
devtools::load_all()
```

```{r message=FALSE}
library(tidyverse)
```

We will use a built-in example dataset, `polyafit_example_data`, to demonstrate
how the package works. The example data is in the form of a matrix, with three
observations and 50 features. In this example, each observation corresponds to
a microbiome sample and each feature corresponds to a bacterial species. The
numbers in the matrix represent the number of times each species was observed
in the sample. Let's print out the example data to see how it looks.

```{r}
polyafit_example_data
```

Looking down each column, we can see that the number of counts per species is
approximately the same for the three samples, except for `species_4`. This
species seems to be over-represented in `sample_3` (150 counts) relative to the
other samples (8 and 9 counts).

Our goals are to measure the overall correspondence in species abundances
between the samples, and then to use this as a yardstick to determine if we
have evidence that a species is over-represented in one of the samples.


### Analysis: no features enriched

To start, we will select the first two samples and run a fit. When the example
data set was generated, these samples were drawn from the same multinomial
distribution. Therefore, we should not detect any features that are enriched in
one sample relative to the other. Furthermore, we expect to observe a high
degree of correspondence between the relative abundances, or in other words, a
low degree of overdispersion relative to the multinomial distribution.

We select the two samples by name, giving a matrix with two rows and 50
columns.

```{r}
example12 <- polyafit_example_data[c("sample_1", "sample_2"),]
```

We find the best-fit parameters of the distribution using the `pfit()`
function.

```{r}
fit12 <- example12 %>%
  pfit()
```

We can call `plot()` on this object to see a nice plot of the data.

```{r warning=FALSE}
fit12 %>%
  plot()
```

Raw p-values for each feature.  All p-values are > 0.05.

```{r}
fit12 %>%
  feature_enrichment() %>%
  filter(p.value < 0.05)
```

The fitted object, `pfit`, has a vector of estimated parameter values for the
Dirichlet-multinomial (DM) distribution, which serves as the mathematical model
for our data. From the estimated parameter values, we can compute an
overdispersion parameter, theta. Theta tells us how much "extra variance" the
estimated distribution has, beyond what we expect from a multinomial. Because
the data were generated from a perfect multinomial distribution, the estimated
overdispersion parameter should be very small. 

```{r}
fit12$theta
```

### Analysis: one outlier

In sample 3, we've adjusted one of the species to increase the number of
counts. We should see this species, `species_4`, to be enriched in our
analysis.

```{r}
example13 <- polyafit_example_data[c("sample_1", "sample_3"),]
```

```{r}
fit13 <- example13 %>%
  pfit()
```

```{r}
fit13 %>%
  plot()
```


```{r}
fit13 %>%
  feature_enrichment() %>%
  filter(p.value < 0.05)
```

Now, we re-calculate the overdispersion parameter, theta. It has increased by
three orders of magnitude.

```{r}
fit13$theta
```


